---
title: "Grammar of Graphics"
subtitle: "ggplot2"
author: "George Fullegar"
date: "17 May 2017"
output: 
  ioslides_presentation:
    widescreen: true
    smaller: true
    incremental: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
require(ggplot2)
require(scales)
require(rvest)
require(stringr)
require(car)
require(data.table)
require(dtplyr)
require(dplyr)
require(rgeos)
require(maptools)
require(rgdal)
require(gdata)
require(knitr)
require(grid)
require(gdata)

theme_set(theme_minimal())


## Convenience "not in" function
"%nin%" <- function(x, y) {
  return( !(x %in% y) )
}

##' Footnote a plot.
##' @title makeFootnote
##' @param footnoteText Defaults to date stamp.
##' @param size size
##' @param colour colour
##' @return plot footnote
##' @author Kieran Healy
##' @export
makeFootnote <- function(footnoteText=
                           format(Sys.time(), "%d %b %Y"),
                         size= .7, colour= grey(.5))
{

  pushViewport(viewport())
  grid.text(label= footnoteText ,
            x = unit(1,"npc") - unit(2, "mm"),
            y= unit(2, "mm"),
            just=c("right", "bottom"),
            gp=gpar(cex= size, col=colour))
  popViewport()
}


### Plot a credit line
##' Credit a plot
##' @title credit
##' @param credit.text Who gets the credit. Defaults to my name
##homepage URL.
##' @return credit
##' @author Kieran Healy
##' @export
##' @param credit.text Text of the credit line
##' @param ...  Other arguments to makeFootnote
credit <- function(credit.text="Kieran Healy, http://kieranhealy.org\n", ...) {
  return(makeFootnote(credit.text, ...))
}


## credit() and makeFootnote() functions can be found here:
## https://github.com/kjhealy/kjhutils/blob/master/R/utility.R




data <- fread("data\\uk-election-results-2015.csv")

#
data <- data %>% 
          group_by(Constituency) %>% 
          mutate(Total.Votes.Cast=sum(Votes), Rank=row_number(desc(Votes)))

data$Party[data$Candidate == "John Bercow"] <- "Conservative"
## Code the Speaker of the House as Conservative

data$Party <- as.factor(data$Party)
##  data$Party <- reorder.factor(data$Party, new.order=by.seats$Party)


by.gb.mps <- data %>% group_by(Constituency) %>%
  filter(Votes==max(Votes) & Region!="Northern Ireland" )  %>%
  ungroup() %>% arrange(desc(Vote.Share))  %>% data.table(.)

by.seats <- data %>% group_by(Constituency) %>% filter(Votes==max(Votes)) %>%
  group_by(Party) %>% tally() %>% arrange(desc(n)) %>%
  data.table(.)

all.parties <- data %>% group_by(Party.all) %>% tally() %>% arrange(desc(n)) %>% data.table()

by.mps <- data %>% group_by(Constituency) %>% filter(Votes==max(Votes))  %>%
  ungroup() %>% arrange(desc(Vote.Share)) %>% data.table(.)
by.mps$Party <- factor(by.mps$Party)

pal <- function(col, border = "light gray")
{
  n <- length(col)
  plot(0, 0, type="n", xlim = c(0, 1), ylim = c(0, 1), axes=FALSE, xlab = "", ylab = "")
  rect(0:(n-1)/n, 0, 1:n/n, 1, col = col, border = border)
}


uk.colours <- data.table(Party=levels(by.mps$Party),
                        party.colour=c(
                          "#1577C7", # Conservative
                          "#BC1D40", # DUP
                          "#7EC031", # Greens
                          "#999999", # Independent
                          "#E8251F", # Labour
                          "#FA8324", # Lim-Dems
 ##                         "#000000", # Other
                          "#559D37", # Plaid Cymru
                          "#EAC135", # SNP
                          "#126140", # Sinn Fein
                          "#6AA769", # SDLP
                          "#6E3485", # UKIP
                          "#233160" # UUP
                        ),
                        stringsAsFactors = FALSE)


pal(uk.colours$party.colour)

not.gb <- c("Democratic Unionist Party", "Sinn Fein", "Social Democratic & Labour Party",
            "Ulster Unionist Party")
gb.colours <- uk.colours %>% filter(Party %nin% not.gb)

pc.look <- function(parties){
    x <- match(parties, uk.colours$Party)
    colours <- uk.colours$party.colour[x]
    return(colours)
}

    constituency_names.df <- fread("data/constituency_names.csv", header=TRUE, stringsAsFactors=FALSE)

    
```

## ggplot2{.build}

Grammar of graphics

- **data**
- **aes**thetic mappings
- **geom**etric objects
- **scale**s
- **facet**s
- *stat*s
- *coord*inate system


## Look at the data {.build}
```{r echo = TRUE}
str(by.mps)
```

## Quick bar chart
```{r echo = TRUE}
p2 <- qplot(Party,data = by.mps,geom="bar")
p2
```

## Tidy up
can't read the labels, so
```{r echo = TRUE}
p2 <- qplot(Party,data = by.mps,geom="bar")
p2 <- p2 + theme(axis.text.x = element_text(angle = 30, hjust = 1))
p2 <- p2 + labs(x="Winning Party", y="Number of Seats")
```

## Tidy up
```{r echo = FALSE}
p2 
```


## Basic Scatter Plot {.build}

```{r echo = TRUE}
p <- ggplot(by.mps, aes(x=Total.Votes.Cast, y=Vote.Share))
p <- p + geom_point() 
p
```

## Add Colour Formatting {.build}

```{r echo = TRUE}
p <- ggplot(by.mps, aes(x=Total.Votes.Cast, y=Vote.Share, colour=Party))
p <- p + geom_point() 
p
```

## Assign Party Colours to Data {.build}

```{r uk.colours, echo = TRUE}
uk.colours
```

## Assign Party Colours to Data {.build}

```{r echo = TRUE}
p <- p + scale_colour_manual(values=uk.colours$party.colour)
p
```

## Add a facet {.build}

```{r echo = TRUE}
p <- p + facet_grid(.~Region)
p <- p + labs(x="Votes Cast in Constituency",
            y="Vote Share of Winner", colour="Winner's Party")

```
## Add a facet {.build}

```{r echo = FALSE}
p <- ggplot(by.mps, aes(x=Total.Votes.Cast, y=Vote.Share, colour=Party))
p <- p + geom_point() 
p <- p + scale_colour_manual(values=uk.colours$party.colour)
p <- p + facet_grid(.~Region)
p <- p + labs(x="Votes Cast in Constituency",
            y="Vote Share of Winner", colour="Winner's Party")
p
```

## Put it all together and tidy up the axis {.build}

```{r echo = TRUE}
p <- ggplot(by.mps, aes(x=Total.Votes.Cast/1000, y=Vote.Share/100, colour=Party))
p <- p + geom_point() 
p <- p + scale_colour_manual(values=uk.colours$party.colour)
p <- p + facet_grid(.~Region)
p <- p + labs(x="Votes Cast in Constituency (Thousands)")
p <- p + labs(y="Vote Share of Winner", colour="Winner's Party")
p <- p + scale_x_continuous(labels=comma)
p <- p + scale_y_continuous(labels=percent)

```
## Put it all together and tidy up the axis {.build}

```{r echo = FALSE}
p <- ggplot(by.mps, aes(x=Total.Votes.Cast/1000, y=Vote.Share/100, colour=Party))
p <- p + geom_point() 
p <- p + scale_colour_manual(values=uk.colours$party.colour)
p <- p + facet_grid(.~Region)
p <- p + labs(x="Votes Cast in Constituency (Thousands)")
p <- p + labs(y="Vote Share of Winner", colour="Winner's Party")
p <- p + scale_x_continuous(labels=comma)
p <- p + scale_y_continuous(labels=percent)
p
```


## Put it all together and tidy up the axis {.build}
replacing

```{r echo = TRUE}
p <- p + facet_grid(.~Region)
```

with

```{r echo = TRUE}
p <- p + facet_grid(Region~.)
```

produces:

## Put it all together and tidy up the axis {.build}
```{r echo = FALSE}

p <- ggplot(by.mps, aes(x=Total.Votes.Cast/1000, y=Vote.Share/100, colour=Party))
p <- p + geom_point() 
p <- p + scale_colour_manual(values=uk.colours$party.colour)
p <- p + facet_grid(Region~.)
p <- p + labs(x="Votes Cast in Constituency (Thousands)")
p <- p + labs(y="Vote Share of Winner", colour="Winner's Party")
p <- p + scale_x_continuous(labels=comma)
p <- p + scale_y_continuous(labels=percent)
p
```


##	A few examples
```{r echo = TRUE}
p <- ggplot(by.mps, aes(x=Total.Votes.Cast, y=Vote.Share/100, colour=Party))
p <- p +  geom_point() + scale_colour_manual(values=uk.colours$party.colour) +
        scale_x_continuous(labels=comma) + scale_y_continuous(labels=percent)
p
```

##	A few examples

```{r echo = TRUE}
safest.seats <- by.mps %>% filter(Vote.Share>63) %>% select(Constituency, Vote.Share, Party) %>%
    data.table()

p <- ggplot(safest.seats, aes(x=reorder(as.character(Constituency), Vote.Share, order=TRUE),
                              y=Vote.Share,
                              colour=Party))
p <- p + geom_point(size=3) + coord_flip() + scale_colour_manual(values=
		pc.look(c("Conservative", "Labour", "Other"))) + labs(x="", y="Winning Candidate's Vote Share") + 
		ggtitle("Safest Seats") + theme(legend.position="top")
```

##	A few examples

```{r echo = FALSE}
p
```

##  A few examples

```{r echo = FALSE}

uk.map <- readOGR("maps/topo_wpc.json", "wpc",verbose = FALSE)
## The name field didn't get imported properly for some reason.
ind <- match(uk.map@data$id, constituency_names.df$id)
uk.map@data$name <- constituency_names.df$Constituency[ind]

constituencies.map <- data.frame(id=0:(length(uk.map@data$name)-1),
                       Constituency=as.character(uk.map@data$name))

uk.map.df <- fortify(uk.map)

uk.map.df <- merge(uk.map.df, constituencies.map, by="id")

## Now we have a map of all the constituencys and winners
uk.map.df <- merge(uk.map.df, by.mps, by="Constituency")

uk.map.df$Party <- as.factor(uk.map.df$Party)

uk.map.df$Party <- droplevels(uk.map.df$Party)

uk.map.df$Party <- reorder.factor(uk.map.df$Party,new.order=as.character(by.seats$Party), order=TRUE)


```

```{r echo = TRUE}
### Make the maps
p <- ggplot(data=uk.map.df, aes(x=long, y=lat, group=group))

p <- p + geom_map(data = uk.map.df, map = uk.map.df,
                   aes(map_id=id, group=group, fill=Party),
                   colour="white", size=0.2)

p <- p + geom_map(data=subset(uk.map.df, Constituency=="York Central"),
                    map=subset(uk.map.df, Constituency=="York Central"),
                    aes(map_id=id, group=group, fill=Party),
                    colour="white", size=0.2)


p <- p + coord_map(projection="albers", at0 = 51, lat1 = 0) + labs(x=NULL, y=NULL, fill="") +
    theme(panel.grid=element_blank(),
          axis.ticks=element_blank(),
          panel.border=element_blank(),
          axis.text=element_blank())

p <- p + scale_fill_manual(values=gb.colours$party.colour,limits=gb.colours$Party)

```

##  A few examples

```{r echo = FALSE}
p

```


## Credits {.build}

The code to source the data is heavily based on the work done for [this blogpost](https://www.r-bloggers.com/who-came-second-in-the-uk-election/ "Who Came Second in the UK Election? Kieran Healy") by Kieran Healy. The data is available on [their github (https://github.com/kjhealy/)](https://github.com/kjhealy/uk-elections/tree/master/data).

The full presentation is available on [my githhub (https://github.com/JorjBeardson)](https://github.com/JorjBeardson)
